To Batch or Not to Batch: Test-Ordering Variability in the ED

### 2024-09-09 Batching Draft 8/Batching-Draft-8.tex
getting going

---
title: "To Batch or Not to Batch: Test-Ordering Variability in the ED"
authors:
  - name: Jacob Jameson
    department:  Interfaculty Initiative in Health Policy
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Soroush Saghafian
    department:  Harvard Kennedy School of Government
    affiliation: Harvard University
    location: Cambridge, MA 
  - name: Robert Huckman
    department:  Harvard Business School
    affiliation: Harvard University
    location: Boston, MA 
  - name: Nicole Hodgson
    department:  Emergency Department
    affiliation: Mayo Clinic
    location: Phoenix, AZ
abstract: |

keywords:

bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{tabularx}
  - \usepackage{graphicx}
  - \usepackage{tikz}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
  - \usepackage{caption}
  - \usepackage{adjustbox}
  - \usepackage{multirow}
---

# Introduction

Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization. Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events. Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality [@bernstein2009].

One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests [@naseim2015]. A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests. These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially. This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing. Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results [@koch2018]. Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier. Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged. Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements [@saghafian2015], however, research surrounding test ordering strategies to improve the patient flow processes remains limited.

@@

# Literature Review

# Theoretical Background and Hypotheses

Emergency departments (EDs) are complex, time-sensitive environments where physicians must make rapid decisions under uncertainty [@asplin2003model; @jarvis2016ed]. Our theoretical model aims to capture the nuanced dynamics of diagnostic decision-making in this high-stakes setting, with a particular focus on the practice of batch ordering imaging tests.

## Diagnostic Decision-Making and Test Ordering in the ED

In the ED, physicians face the challenge of diagnosing and treating patients efficiently while maintaining high-quality care [@croskerry2013cognitive; @johnson2019decision]. Upon a patient's arrival, the physician must decide on a diagnostic strategy, which often involves ordering imaging tests. This decision-making process is influenced by various factors, including the physician's experience, the patient's presenting symptoms, and the ED's current state (e.g., crowding levels) [@cooke2013overview; @bernstein2009effect; @morley2018emergency].

The key distinction in our study is between batch ordering and sequential ordering of imaging tests. In batch ordering, a physician orders multiple tests simultaneously, while in sequential ordering, tests are ordered one at a time, with each subsequent test decision informed by the results of the previous tests. It's crucial to understand that batch ordering does not mean tests are performed simultaneously. Instead, it means that patients enter the queues for multiple tests at once. This distinction is important when considering the implications of batch ordering on ED operations and patient flow [@green2006managing; @batt2019impact].

## Queuing Theory and ED Operations

When a physician batch orders tests, the patient enters multiple queues simultaneously. This can be modeled using parallel queuing systems in operations research [@dai2010load; @zychlinski2019discrete]. While this might seem to expedite the process, it can lead to inefficiencies in resource allocation, as multiple departments (e.g., radiology, laboratory) must allocate resources to a single patient simultaneously, potentially leading to suboptimal resource utilization [@shi2016models; @wang2019data].

Moreover, batch ordering creates an information lag where results from one test cannot inform the decision to proceed with subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis]. This approach can also create blocking effects, where a patient waiting for multiple tests may block other patients who need only one of those tests, potentially increasing overall waiting times [@armony2015heterogeneous; @song2020managing].

## Reasons for Batch Ordering and Theoretical Implications

Physicians may choose to batch order for several reasons. They may believe it will speed up the diagnostic process [@kocher2012effect; @rosen2018quality], or they may be practicing defensive medicine to avoid missing critical diagnoses [@studdert2005defensive; @kanzaria2015emergency]. Batch ordering may also reduce the cognitive load required to make multiple sequential decisions [@croskerry2013cognitive; @nibbelink2018emergency], which can be particularly appealing in busy EDs where physicians are under time pressure [@bernstein2009effect; @morley2018emergency].

From a theoretical perspective, batch ordering presents several implications. It forgoes the opportunity for value of information considerations, where each test result could inform subsequent decisions [@smith2013impact; @akhlaghpour2020value]. Sequential ordering preserves the option to stop testing once a diagnosis is reached, aligning with real options theory in decision-making under uncertainty [@dixit1994investment; @garg2018value]. While batch ordering might seem efficient for individual patients, it could lead to systemic inefficiencies in resource utilization and patient flow [@green2006managing; @batt2019impact].

## Hypotheses Development

@@

### Impact on Resource Utilization

*Hypothesis 1: Batch ordering of diagnostic tests will lead to increased overall test utilization compared to sequential ordering.* This hypothesis is grounded in the idea that batch ordering bypasses the opportunity to use information from initial tests to inform the necessity of subsequent tests, potentially leading to unnecessary testing [@welch2011overdiagnosed; @o2018overdiagnosis].

### Effect on Length of Stay

*Hypothesis 2: Batch ordering will have a significant impact on patient length of stay (LOS) in the ED, but the direction of this effect is ambiguous.* While batch ordering might reduce LOS by eliminating waiting times between test decisions [@kocher2012effect; @rosen2018quality], it could also increase LOS due to potential queuing inefficiencies and the time required to perform and interpret additional tests [@forster2003effect; @armony2015heterogeneous; @song2020managing].

### Quality of Care and Return Visits

*Hypothesis 3: Batch ordering will influence the rate of 72-hour return visits with admission, but the direction of this effect is uncertain.* Comprehensive initial testing might reduce missed diagnoses and thus return visits [@pines2010national; @sabbatini2016overzealous]. However, it could also lead to false positives and unnecessary follow-ups, potentially increasing return visits [@welch2011overdiagnosed; @o2018overdiagnosis].

### Hypothesis 4: Heterogeneity Across Presenting Complaints

*Hypothesis 4: The effects of batch ordering on resource utilization, length of stay, and quality of care will vary significantly across different presenting complaints.* This hypothesis is based on the understanding that different clinical presentations have varying levels of diagnostic uncertainty and risk [@cooke2013overview; @johnson2019decision], which may influence the appropriateness and impact of batch ordering.

@@

# Data and Methods

@@

## Mayo Clinic Data

Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a tertiary care hospital without obstetrical services, an inpatient pediatrics unit, or a trauma designation. During the study period from October 6, 2018, through December 31, 2019, the ED recorded 48,854 visits per year, managed across 26 treatment rooms and up to 9 hallway spaces. The department is exclusively staffed by board-eligible or board-certified emergency physicians (EPs), with rotating residents overseeing about 10% of patient volume. The data is summarized in Table \ref{tab:summary_stats}.

@@

A key feature of the Mayo Clinic ED is its rotational patient assignment system, in which patients arriving at the Mayo Clinic ED are randomly assigned to physicians via a rotational patient assignment algorithm @traub2016emergency, which removes potential selection bias concerns for our analyses. In essence, barring arrival time and shift-level variation, the physician-to-patient matching can be deemed random. Table \ref{tab:wald_test} displays that patient encounters (regarding chief complaints and emergency severity) are equitably distributed across physicians within our study's cohort.

@@

## Sample Construction

Our research design focuses on adults who visit the Mayo Clinic of Arizona ED. We observe $48,854$ such visits during the study period. To improve power, we drop encounters with rare chief complaints ($<1000$ total encounters of this kind) and complaints where a batch order occurs less than $5$ percent of the time. Since batch orders are rare for these cases, our physician batch tendency instrument could suffer from a weak instrument problem if we included them. Examples of complaints dropped include Upper Respiratory Symptoms and Urinary Complaints. Excluding these conditions does not introduce selection bias unless physician test batching tendency is orthogonal to physician diagnosing behavior. While this assumption may be violated if we were to use a very detailed level of chief complaint information upon which to base our exclusion criterion, it is plausibly satisfied when using broad complaint categories as we do. In order to estimate a precise measure of physician-level batch tendency, we further restrict our sample to the $15,821$ encounters involving full-time physicians who treat over 500 ED cases per year.

## Variable Definitions

Our explanatory variable in the IV analysis, $Batched_i$, is an indicator for whether patient $i$ has their tests batch-ordered at their ED encounter. While the patient could decide not to undergo the tests ordered by the physician, this is rare in practice. We define "batching" in line with standard emergency medicine practices. Batching occurs when a physician simultaneously orders a comprehensive set of diagnostic tests, typically covering a broad range of potential diagnoses. This contrasts with sequential ordering, where tests are ordered sequentially based on the information obtained from each test as needed.

We operationalize batching as occurring when multiple diagnostic imaging tests are ordered within a 5-minute window. In Section [...], sensitivity analyses on this cutoff point showed that our results are robust to this definition. Each imaging test (e.g., X-ray, CT scan, Ultrasound) is considered a separate, distinct test for our study. Therefore, a batch in our study consists of two or more distinct imaging tests.

@@

*(a) ED length of stay (LOS).*-- ED-LOS is a critical measure of efficiency and patient throughput in emergency care settings. It is defined as the duration from a patient's arrival to the ED until their departure, whether by discharge or admission to the hospital.

*(b) Number of Unique Imaging Tests*-- Resource utilization in the ED typically refers to the extent of medical services and interventions a patient receives. In this study, we quantify resource utilization by the total number of distinct imaging tests ordered per patient during their ED stay. This encompasses both initial and subsequent tests.

*(c) 72 Hour Return with Admission*-- This is a binary variable indicating whether the patient returns to the ED within 72 hours of their initial visit and is admitted to the hospital. This outcome is used to assess the quality of care provided during the initial ED visit.

## Empirical Strategy

Our empirical strategy closely follows the literature that relies on quasi-random assignment of agents to cases, often referred to as the “judges design.” Papers in this literature typically exploit variation in the sentencing leniency of judges who work in the same court. Similarly, we explore batching variation across physicians who work in the same emergency department. In its reduced form, under the assumption of quasi-random assignment, this approach allows researchers to identify the causal effect of being assigned to different types of physicians. Under additional assumptions, an instrumental variable approach identifies the causal effect of a given medical decision. We employ both approaches and lay out their details in the next subsections.

### Batch Tendency Construction

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

@@

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their imaging tests batch ordered on encounter that took place on date $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. Chief complaint by severity fixed effects, $\alpha_{complaint}$, were also included to increase precision. Finally, a binary variable for whether or not laboratory tests were ordered, $\alpha_{lab}$, was included to account for the complexity of the case. As stated earlier, these controls are more than what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the tendency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

@@

where $\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order imaging tests. Table \ref{table:first_stage} presents the ``first stage" in a regression table: being assigned to a $10$pp higher batch-tendency physician is associated with a $6$pp increase in the likelihood of having tests batch-ordered in the ED. The F-statistic is 94 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the tendency instrument, while the first stage is calculated using the baseline sample only, which excludes the rare complaints.

@@

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour return with admission, and number of imaging tests ordered for the patient, and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, length of stay, and return with admission likelihood. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$. We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

## Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, since assignment of patients to ED physicians is random, conditional on seasonality and shift (``conditional independence"). The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal effect of having diagnostic tests batch ordered, relies on three additional assumptions: relevance, exclusion, and monotonicity. We reported a strong first stage (i.e., relevance) at the end of the previous Section. The exclusion restriction requires that the instrument must influence the outcome of interest only through its effect on test batching. This is perhaps our strongest assumption and is at its core, untestable. However, several features of the ED setting suggest that such violation may likely only have a small impact and may be less concerning than in other health care settings. First, unlike in primary care settings, where the patient and primary care provider have many repeat encounters, the scope of what the emergency physician can do to impact medium-term outcomes is limited and well-observed by the researcher. Second, any violation of the exclusion restriction needs to directly affect the specific outcome of interest. The channel by which ED physicians can influence length of stay relative outcomes is likely through testing and diagnosis. Nevertheless, we take this assumption seriously and perform a placebo check in Section [...] as well as various robustness checks in Section [...].

>>>>

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{To Batch or Not to Batch: Test-Ordering Variability in the ED}

\author{
    Jacob Jameson
   \\
    Interfaculty Initiative in Health Policy \\
    Harvard University \\
  Cambridge, MA \\
  \texttt{} \\
   \And
    Soroush Saghafian
   \\
    Harvard Kennedy School of Government \\
    Harvard University \\
  Cambridge, MA \\
  \texttt{} \\
   \And
    Robert Huckman
   \\
    Harvard Business School \\
    Harvard University \\
  Boston, MA \\
  \texttt{} \\
   \And
    Nicole Hodgson
   \\
    Emergency Department \\
    Mayo Clinic \\
  Phoenix, AZ \\
  \texttt{} \\
  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}
\usepackage{caption}
\usepackage{adjustbox}
\usepackage{multirow}
\begin{document}
\maketitle

\begin{abstract}

\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Healthcare delivery, particularly in the emergency department (ED), is a
delicate balance that involves ensuring optimal patient outcomes while
optimizing resource utilization. Achieving these twin goals requires
timely and accurate diagnosis, which in turn enables prompt and
appropriate treatment, consequently improving patient prognosis and
reducing the likelihood of adverse events. Furthermore, efficient
patient discharge from the ED can help alleviate overcrowding, a severe
issue with potential consequences including higher complication rates
and increased mortality (S. Bernstein et al. 2009).

One important factor that can impact the speed and effectiveness of
diagnosis in the ED is the availability and performance of diagnostic
tests (Balogh, Miller, and Ball 2015). A variety of diagnostic tests are
used in the ED, including laboratory tests, imaging studies, and
specialized tests. These tests can provide valuable information about a
patient's condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in
the ED should batch order diagnostic tests or order them sequentially.
This decision essentially represents a tradeoff between reducing patient
length of stay and risk of over-testing. Over-testing, or performing
unnecessary tests, can lead to increased costs, unnecessary patient
anxiety, and potential harm from follow-up of false-positive results
(Koch et al. 2018). Conversely, keeping a patient for an extended time
to perform all possible tests could lead to ED overcrowding, an issue
associated with severe consequences, as mentioned earlier. Instead, what
is needed is a reasonable balance between the number of diagnostic tests
performed and the total time the patient is kept in the ED before either
being admitted or discharged. Several studies have demonstrated that
optimizing the ED patient flow process can result in significant
improvements (Saghafian, Austin, and Traub 2015), however, research
surrounding test ordering strategies to improve the patient flow
processes remains limited.

@@

\hypertarget{literature-review}{%
\section{Literature Review}\label{literature-review}}

\hypertarget{theoretical-background-and-hypotheses}{%
\section{Theoretical Background and
Hypotheses}\label{theoretical-background-and-hypotheses}}

Emergency departments (EDs) are complex, time-sensitive environments
where physicians must make rapid decisions under uncertainty (Asplin et
al. 2003; Jarvis 2016). Our theoretical model aims to capture the
nuanced dynamics of diagnostic decision-making in this high-stakes
setting, with a particular focus on the practice of batch ordering
imaging tests.

\hypertarget{diagnostic-decision-making-and-test-ordering-in-the-ed}{%
\subsection{Diagnostic Decision-Making and Test Ordering in the
ED}\label{diagnostic-decision-making-and-test-ordering-in-the-ed}}

In the ED, physicians face the challenge of diagnosing and treating
patients efficiently while maintaining high-quality care (Croskerry
2013; Johnson 2019). Upon a patient's arrival, the physician must decide
on a diagnostic strategy, which often involves ordering imaging tests.
This decision-making process is influenced by various factors, including
the physician's experience, the patient's presenting symptoms, and the
ED's current state (e.g., crowding levels) (Cooke et al. 2013; S. L.
Bernstein et al. 2009; Morley et al. 2018).

The key distinction in our study is between batch ordering and
sequential ordering of imaging tests. In batch ordering, a physician
orders multiple tests simultaneously, while in sequential ordering,
tests are ordered one at a time, with each subsequent test decision
informed by the results of the previous tests. It's crucial to
understand that batch ordering does not mean tests are performed
simultaneously. Instead, it means that patients enter the queues for
multiple tests at once. This distinction is important when considering
the implications of batch ordering on ED operations and patient flow
(Green 2006; Batt and Terwiesch 2019).

\hypertarget{queuing-theory-and-ed-operations}{%
\subsection{Queuing Theory and ED
Operations}\label{queuing-theory-and-ed-operations}}

When a physician batch orders tests, the patient enters multiple queues
simultaneously. This can be modeled using parallel queuing systems in
operations research (Dai and He 2010; Zychlinski et al. 2019). While
this might seem to expedite the process, it can lead to inefficiencies
in resource allocation, as multiple departments (e.g., radiology,
laboratory) must allocate resources to a single patient simultaneously,
potentially leading to suboptimal resource utilization (Shi et al. 2016;
Wang et al. 2019).

Moreover, batch ordering creates an information lag where results from
one test cannot inform the decision to proceed with subsequent tests,
potentially leading to unnecessary testing (Welch, Schwartz, and
Woloshin 2011; O'Connor and Nichol 2018). This approach can also create
blocking effects, where a patient waiting for multiple tests may block
other patients who need only one of those tests, potentially increasing
overall waiting times (Armony et al. 2015; Song et al. 2020).

\hypertarget{reasons-for-batch-ordering-and-theoretical-implications}{%
\subsection{Reasons for Batch Ordering and Theoretical
Implications}\label{reasons-for-batch-ordering-and-theoretical-implications}}

Physicians may choose to batch order for several reasons. They may
believe it will speed up the diagnostic process (Kocher et al. 2012;
Rosen and Seda 2018), or they may be practicing defensive medicine to
avoid missing critical diagnoses (Studdert et al. 2005; Kanzaria et al.
2015). Batch ordering may also reduce the cognitive load required to
make multiple sequential decisions (Croskerry 2013; Nibbelink and Brewer
2018), which can be particularly appealing in busy EDs where physicians
are under time pressure (S. L. Bernstein et al. 2009; Morley et al.
2018).

From a theoretical perspective, batch ordering presents several
implications. It forgoes the opportunity for value of information
considerations, where each test result could inform subsequent decisions
(Smith et al. 2013; Akhlaghpour 2020). Sequential ordering preserves the
option to stop testing once a diagnosis is reached, aligning with real
options theory in decision-making under uncertainty (Dixit and Pindyck
1994; Garg, Wang, and Valerdi 2018). While batch ordering might seem
efficient for individual patients, it could lead to systemic
inefficiencies in resource utilization and patient flow (Green 2006;
Batt and Terwiesch 2019).

\hypertarget{hypotheses-development}{%
\subsection{Hypotheses Development}\label{hypotheses-development}}

@@

\hypertarget{impact-on-resource-utilization}{%
\subsubsection{Impact on Resource
Utilization}\label{impact-on-resource-utilization}}

\emph{Hypothesis 1: Batch ordering of diagnostic tests will lead to
increased overall test utilization compared to sequential ordering.}
This hypothesis is grounded in the idea that batch ordering bypasses the
opportunity to use information from initial tests to inform the
necessity of subsequent tests, potentially leading to unnecessary
testing (Welch, Schwartz, and Woloshin 2011; O'Connor and Nichol 2018).

\hypertarget{effect-on-length-of-stay}{%
\subsubsection{Effect on Length of
Stay}\label{effect-on-length-of-stay}}

\emph{Hypothesis 2: Batch ordering will have a significant impact on
patient length of stay (LOS) in the ED, but the direction of this effect
is ambiguous.} While batch ordering might reduce LOS by eliminating
waiting times between test decisions (Kocher et al. 2012; Rosen and Seda
2018), it could also increase LOS due to potential queuing
inefficiencies and the time required to perform and interpret additional
tests (Forster et al. 2003; Armony et al. 2015; Song et al. 2020).

\hypertarget{quality-of-care-and-return-visits}{%
\subsubsection{Quality of Care and Return
Visits}\label{quality-of-care-and-return-visits}}

\emph{Hypothesis 3: Batch ordering will influence the rate of 72-hour
return visits with admission, but the direction of this effect is
uncertain.} Comprehensive initial testing might reduce missed diagnoses
and thus return visits (Pines et al. 2010; Sabbatini et al. 2016).
However, it could also lead to false positives and unnecessary
follow-ups, potentially increasing return visits (Welch, Schwartz, and
Woloshin 2011; O'Connor and Nichol 2018).

\hypertarget{hypothesis-4-heterogeneity-across-presenting-complaints}{%
\subsubsection{Hypothesis 4: Heterogeneity Across Presenting
Complaints}\label{hypothesis-4-heterogeneity-across-presenting-complaints}}

\emph{Hypothesis 4: The effects of batch ordering on resource
utilization, length of stay, and quality of care will vary significantly
across different presenting complaints.} This hypothesis is based on the
understanding that different clinical presentations have varying levels
of diagnostic uncertainty and risk (Cooke et al. 2013; Johnson 2019),
which may influence the appropriateness and impact of batch ordering.

@@

\hypertarget{data-and-methods}{%
\section{Data and Methods}\label{data-and-methods}}

@@

\hypertarget{mayo-clinic-data}{%
\subsection{Mayo Clinic Data}\label{mayo-clinic-data}}

Our primary dataset comes from the ED of the Mayo Clinic of Arizona, a
tertiary care hospital without obstetrical services, an inpatient
pediatrics unit, or a trauma designation. During the study period from
October 6, 2018, through December 31, 2019, the ED recorded 48,854
visits per year, managed across 26 treatment rooms and up to 9 hallway
spaces. The department is exclusively staffed by board-eligible or
board-certified emergency physicians (EPs), with rotating residents
overseeing about 10\% of patient volume. The data is summarized in Table
\ref{tab:summary_stats}.

@@

A key feature of the Mayo Clinic ED is its rotational patient assignment
system, in which patients arriving at the Mayo Clinic ED are randomly
assigned to physicians via a rotational patient assignment algorithm
Traub et al. (2016), which removes potential selection bias concerns for
our analyses. In essence, barring arrival time and shift-level
variation, the physician-to-patient matching can be deemed random. Table
\ref{tab:wald_test} displays that patient encounters (regarding chief
complaints and emergency severity) are equitably distributed across
physicians within our study's cohort.

@@

\hypertarget{sample-construction}{%
\subsection{Sample Construction}\label{sample-construction}}

Our research design focuses on adults who visit the Mayo Clinic of
Arizona ED. We observe \(48,854\) such visits during the study period.
To improve power, we drop encounters with rare chief complaints
(\(<1000\) total encounters of this kind) and complaints where a batch
order occurs less than \(5\) percent of the time. Since batch orders are
rare for these cases, our physician batch tendency instrument could
suffer from a weak instrument problem if we included them. Examples of
complaints dropped include Upper Respiratory Symptoms and Urinary
Complaints. Excluding these conditions does not introduce selection bias
unless physician test batching tendency is orthogonal to physician
diagnosing behavior. While this assumption may be violated if we were to
use a very detailed level of chief complaint information upon which to
base our exclusion criterion, it is plausibly satisfied when using broad
complaint categories as we do. In order to estimate a precise measure of
physician-level batch tendency, we further restrict our sample to the
\(15,821\) encounters involving full-time physicians who treat over 500
ED cases per year.

\hypertarget{variable-definitions}{%
\subsection{Variable Definitions}\label{variable-definitions}}

Our explanatory variable in the IV analysis, \(Batched_i\), is an
indicator for whether patient \(i\) has their tests batch-ordered at
their ED encounter. While the patient could decide not to undergo the
tests ordered by the physician, this is rare in practice. We define
``batching'' in line with standard emergency medicine practices.
Batching occurs when a physician simultaneously orders a comprehensive
set of diagnostic tests, typically covering a broad range of potential
diagnoses. This contrasts with sequential ordering, where tests are
ordered sequentially based on the information obtained from each test as
needed.

We operationalize batching as occurring when multiple diagnostic imaging
tests are ordered within a 5-minute window. In Section {[}\ldots{]},
sensitivity analyses on this cutoff point showed that our results are
robust to this definition. Each imaging test (e.g., X-ray, CT scan,
Ultrasound) is considered a separate, distinct test for our study.
Therefore, a batch in our study consists of two or more distinct imaging
tests.

@@

\emph{(a) ED length of stay (LOS).}-- ED-LOS is a critical measure of
efficiency and patient throughput in emergency care settings. It is
defined as the duration from a patient's arrival to the ED until their
departure, whether by discharge or admission to the hospital.

\emph{(b) Number of Unique Imaging Tests}-- Resource utilization in the
ED typically refers to the extent of medical services and interventions
a patient receives. In this study, we quantify resource utilization by
the total number of distinct imaging tests ordered per patient during
their ED stay. This encompasses both initial and subsequent tests.

\emph{(c) 72 Hour Return with Admission}-- This is a binary variable
indicating whether the patient returns to the ED within 72 hours of
their initial visit and is admitted to the hospital. This outcome is
used to assess the quality of care provided during the initial ED visit.

\hypertarget{empirical-strategy}{%
\subsection{Empirical Strategy}\label{empirical-strategy}}

Our empirical strategy closely follows the literature that relies on
quasi-random assignment of agents to cases, often referred to as the
``judges design.'' Papers in this literature typically exploit variation
in the sentencing leniency of judges who work in the same court.
Similarly, we explore batching variation across physicians who work in
the same emergency department. In its reduced form, under the assumption
of quasi-random assignment, this approach allows researchers to identify
the causal effect of being assigned to different types of physicians.
Under additional assumptions, an instrumental variable approach
identifies the causal effect of a given medical decision. We employ both
approaches and lay out their details in the next subsections.

\hypertarget{batch-tendency-construction}{%
\subsubsection{Batch Tendency
Construction}\label{batch-tendency-construction}}

To measure physician batch tendency, we use the physician's residualized
leave-out average batch rate. This measure is derived from two steps
following the approaches taken by Doyle et al. (2015), Dobbie, Goldin,
and Yang (2018), and Eichmeyer and Zhang (2022). First, we obtain
residuals from a regression model, which includes all ED encounters in
our sample period.

@@

Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
\(i\) had their imaging tests batch ordered on encounter that took place
on date \(t\). Fixed effects include year-month fixed effects,
\(\alpha_{ym}\), to control for time and seasonal variation in batching,
such as hospital-specific policies (e.g.~initiatives to eliminate excess
testing) or seasonality in ED visits. We also control for
``shift-level'' variations that include both physician scheduling and
patient arrival with day of week-time of day fixed effects,
\(\alpha_{dt}\). Chief complaint by severity fixed effects,
\(\alpha_{complaint}\), were also included to increase precision.
Finally, a binary variable for whether or not laboratory tests were
ordered, \(\alpha_{lab}\), was included to account for the complexity of
the case. As stated earlier, these controls are more than what is
required for our quasi-random assignment assumption. Under the
assumption that we have captured the observables under which
quasi-random assignment occurs in the ED, the unexplained variation--
the physician's contribution-- resides in the error term,
\(\varepsilon_{i,t}\).

In step two, the tendency measure for patient \(i\) seen by physician
\(j\) is computed as the average residual across all other patients seen
by the physician that year:

@@

where \(\hat{\varepsilon}_{i'} = \hat{Batch}_{i'} - Batch_{i'}\) is the
residual from equation (1); \(J\) is the set of all ED encounters
treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
number of cases that physician has seen that year, excluding patient
\(i\). This leave-out mean eliminates the mechanical bias that stems
from patient \(i\)'s own case entering into the instrument. The measure
is interpreted as the average (leave-out) batch rate of patient \(i\)'s
physician, relative to other physicians in that hospital-year-month,
hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order imaging tests. Table
\ref{table:first_stage} presents the ``first stage'' in a regression
table: being assigned to a \(10\)pp higher batch-tendency physician is
associated with a \(6\)pp increase in the likelihood of having tests
batch-ordered in the ED. The F-statistic is 94 when all controls and
fixed effects are included. The coefficient is greater than one because
all emergency visits are used to construct the tendency instrument,
while the first stage is calculated using the baseline sample only,
which excludes the rare complaints.

@@

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour return with admission, and number of imaging tests ordered for
the patient, and \(X_i\) is the same as in the reduced-form approach.
\(Batched_i\) variable suffers from potential endogeneity concerns. For
example, injury severity may be unobserved and correlated with need to
run multiple tests, length of stay, and return with admission
likelihood. Hence, we instrument \(Batched_i\) with the assigned
physician \(j\)'s underlying tendency to batch,
\(Tendency_{i,j}^{phys}\). We cluster robust standard errors at the
physician level to account for the assignment process of patients to
physicians.

\hypertarget{identifying-assumptions}{%
\subsection{Identifying Assumptions}\label{identifying-assumptions}}

The reduced-form approach delivers an unbiased estimate of the causal
effect of being treated by a higher tendency to batch physician, since
assignment of patients to ED physicians is random, conditional on
seasonality and shift (``conditional independence''). The
residualization in equation (1) controls for more controls than required
to achieve quasi-random assignment; they are included for statistical
precision in measuring physician tendency to batch.

Our instrumental variable approach, which aims to recover the causal
effect of having diagnostic tests batch ordered, relies on three
additional assumptions: relevance, exclusion, and monotonicity. We
reported a strong first stage (i.e., relevance) at the end of the
previous Section. The exclusion restriction requires that the instrument
must influence the outcome of interest only through its effect on test
batching. This is perhaps our strongest assumption and is at its core,
untestable. However, several features of the ED setting suggest that
such violation may likely only have a small impact and may be less
concerning than in other health care settings. First, unlike in primary
care settings, where the patient and primary care provider have many
repeat encounters, the scope of what the emergency physician can do to
impact medium-term outcomes is limited and well-observed by the
researcher. Second, any violation of the exclusion restriction needs to
directly affect the specific outcome of interest. The channel by which
ED physicians can influence length of stay relative outcomes is likely
through testing and diagnosis. Nevertheless, we take this assumption
seriously and perform a placebo check in Section {[}\ldots{]} as well as
various robustness checks in Section {[}\ldots{]}.
